The Design, Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-30 paper.tex
Novo titulo: The Design, Semantics, and Implementation of Ceu: ... Revisao abstract, Intro

% Title portion
\title{The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel}

@@

\begin{abstract}
\CEU is a reactive language based on Esterel that targets constrained embedded 
platforms.
%
Employing the synchronous programming model, it allows for a simple reasoning 
about shared-memory concurrency.
Furthermore, its restricted semantics enables deterministic and memory-safe 
programs.
%
% TODO: some uses?
%
In this work, we propose a formal semantics for \CEU focusing on its particular 
control mechanisms, such as parallel compositions, finalization, and 
stack-based internal events.
%
We also present an implementation with two backends:
one aiming for resource efficiency and interoperability with C, and another for 
dynamic reprogramming over the air.
\end{abstract}

@@

An established alternative to \C in the field of embedded systems is the family 
of reactive synchronous languages \cite{rp.twelve}.
%
Two major styles of synchronous languages have evolved:
in the \emph{control}--\emph{imperative} style, programs are structured with 
control flow primitives, such as parallelism, repetition, and preemption;
in the \emph{dataflow}--\emph{declarative} style, programs can be seen as 
graphs of values, in which a change to a value is propagated through its 
dependencies without explicit programming.
%
Among the control-based languages, Esterel~\cite{esterel.ieee91} influenced a 
number of other embedded languages, such as \emph{Reactive-C}~\cite{rp.rc}, 
\emph{OSM}~\cite{wsn.osm}, \emph{Sync-C}~\cite{rp.synchc}, and 
\emph{PRET-C}~\cite{rp.pretc}.

\CEU is another language targeting embedded systems based on Esterel with some 
novel functionalities:
%
\begin{itemize}
%
\item Stack-based internal events, which provide a limited form of coroutines.
%
\item A static temporal analysis and deterministic execution semantics that 
allows programs to safely share memory.
%
\item A finalization mechanism for safe abortion of lines of execution.
%
\item First-class synchronized timers.
\end{itemize}

\CEU shares limitations with Esterel and synchronous languages in general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis~\cite{rp.hypothesis}, and cannot be 
elegantly implemented.

In this work, we present a formal semantics of a small synchronous kernel that 
represents a subset of \CEU covering these new functionalities.
%
We also present an implementation of \CEU with two backends:
one aiming for efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of constrained wireless sensor networks, developing a number of applications, 
protocols, and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C, and observed a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for ROM and RAM)~\cite{ceu.sensys13}.
%
Considering the \VM backend, a simple application that blinks three LEDs 
periodically occupies less than 100 bytes and can be completely transmitted in 
4 radio messages.
In contrast, without the preinstalled \VM with basic \emph{I/O} functionality, 
a \C-compiled version occupies more than 2 kbytes~\cite{ceu.terra}.

The rest of the paper is organized as follows:
Section~\ref{sec.ceu} gives an overview of \CEU, exposing its fundamental 
differences to Esterel.
Section~\ref{sec.sem} presents a formal semantics for the control primitives of 
\CEU.
Section~\ref{sec.impl} presents the \C and \VM implementation backends.
Section~\ref{sec.related} discusses other synchronous languages targeting 
embedded systems.
Section~\ref{sec.conclusion} concludes the paper.

@@

\section{Overview of \CEU}
\label{sec.ceu}

@@

\CEU, following the Esterel mindset, has a strong imperative flavor, with 
explicit control flow through sequences, loops, parallels, and also 
assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
In addition, \CEU also employs the synchronous model, in which programs advance 
in a sequence of discrete reactions to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
native calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.

@@

TODO: exemplo de implementacao/uso de um componente
- Terra: binding em nesC
- Ceu: header de configuracao
- (analogo a Figura 19)

@@

TODO: falar um pouco do assembly/bytecode, quais as decisoes envolvidas
- (analogo a Figura 18)

@@

>>>>

% Title portion
\title{The Design, Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel}

@@

\begin{abstract}
\CEU is a reactive language based on Esterel that targets constrained embedded 
platforms and ensures safe concurrency by handling threats at compile time, 
rather than at runtime.
%
Based on the synchronous programming model, our design allows for a simple 
reasoning about concurrency that enables compile-time analysis resulting in 
deterministic and memory-safe programs.
%
% TODO: some uses?
%
In this work, we discuss the design of \CEU and propose a formal semantics 
focusing on its particular control mechanisms, such as parallel compositions, 
finalization, and stack-based internal events.
%
We also present an implementation with two backends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
\end{abstract}

@@

An established alternative to \C in the field of embedded systems is the family 
of reactive synchronous languages \cite{rp.twelve}.
%
Two major styles of synchronous languages have evolved:
in the \emph{control}--\emph{imperative} style, programs are structured with 
control flow primitives, such as parallelism, repetition, and preemption;
in the \emph{dataflow}--\emph{declarative} style, programs can be seen as 
graphs of values, in which a change to a value is propagated through its 
dependencies without explicit programming.
%
Considering the control-based languages, Esterel~\cite{esterel.ieee91} was the 
first to appear and influenced a number of embedded languages, such as 
\emph{Reactive-C}~\cite{rp.rc}, \emph{OSM}~\cite{wsn.osm}, 
\emph{Sync-C}~\cite{rp.synchc}, and \emph{PRET-C}~\cite{rp.pretc}.

\CEU is another Esterel-based language targeting embedded systems with novel 
functionalities:
%
\begin{itemize}
%
\item Stack-based execution for internal events, which provide a limited form 
of coroutines.
%
\item A static temporal analysis and deterministic execution semantics that 
allows programs to safely share memory.
%
\item A finalization mechanism for safe abortion of lines of execution.
%
\item First-class synchronized timers.
\end{itemize}

In this work, we discuss the design of \CEU and propose a formal semantics of a 
small synchronous kernel that represents a subset of the language covering 
these new functionalities.
%
We also present an implementation of \CEU with two backends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes, which use the \emph{ATmega328} 8-bit 
microcontroller with 32 Kbytes of FLASH and 2 Kbytes of SRAM.%
\footnote{
Arduino: \url{https://www.arduino.cc/en/Main/arduinoBoardUno}
\\
MICAz: \url{http://www.memsic.com/userfiles/files/Datasheets/WSN/micaz_datasheet-t.pdf}
\\
ATmega328: \url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
Considering the \VM backend, a simple application that blinks three LEDs 
periodically occupies less than 100 bytes and can be completely transmitted in 
4 radio messages.
In contrast, without the preinstalled \VM with basic \emph{I/O} functionality, 
a \C-compiled version occupies more than 2 Kbytes~\cite{ceu.terra}.

The rest of the paper is organized as follows:
Section~\ref{sec.ceu} discusses the design of \CEU, exposing its fundamental 
differences to Esterel.
Section~\ref{sec.sem} presents a formal semantics for the control primitives of 
\CEU.
Section~\ref{sec.impl} presents the \C and \VM implementation backends.
Section~\ref{sec.related} discusses other synchronous languages targeting 
embedded systems.
Section~\ref{sec.conclusion} concludes the paper.

@@

\section{The Design of \CEU}
\label{sec.ceu}

@@

\CEU, following the Esterel mindset, has a strong imperative flavor, with 
explicit control flow through sequences, loops, parallels, and also 
assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
In addition, \CEU also employs the synchronous model, in which programs advance 
in a sequence of discrete reactions to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
native calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with Esterel and synchronous languages in 
general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis~\cite{rp.hypothesis}, and cannot be 
elegantly implemented.

@@

TODO-VM: exemplo de implementacao/uso de um componente
- Terra: binding em nesC
- Ceu: header de configuracao
- (analogo a Figura 19)

@@

TODO-VM: falar um pouco do assembly/bytecode, quais as decisoes envolvidas
- (analogo a Figura 18)

@@

TODO-VM
